---
title: "Week 1 R programming assignment"
output: html_notebook
Name :- Rathnakar Rajanala
---

** load the datasets
```{r}
# loading necessary libraries
library(dplyr)
library(readr)

# load given dataset's
#outpatient visits
outpatient_visits <- fread("OutpatientVisit.csv") 
#patients
patients <- fread("Patient.csv")   
#clinics
clinics <- fread("Clinic.csv")  
#staff
staff <- fread("Staff.csv")  
#mortality
mortality <- fread("Mortality.csv")  
#icd codes
icd_codes <- fread("ICDCodes.csv")  
#disease map
disease_map <- fread("DiseaseMap.csv")              

```


```{r}
#merge patient data with mortality 
patient_mortality <- merge(patients,mortality,by = "PatientID", all.x = TRUE)

#data.table
setDT(patient_mortality)

# Create new column to indicate if patient is deceased(1) or alive(0)
patient_mortality[,IsDeceased := !is.na(DateOfDeath)]

#group by gender and calculate mortality rate
mortality_rate <- patient_mortality[,.(
  Totalpatients = .N,
  DeceasedPatients = sum(IsDeceased ,na.rm = TRUE),
  Mortality_Rate = sum(IsDeceased ,na.rm = TRUE)/ .N *100
),by = Gender] 

# print mortality rate
print(mortality_rate)
```
****INTERPRETATION****

-->Males have a slightly higher death rate than females in this dataset (35.95% vs. 35.11%).



2.Are there patterns in the disease groups across gender. For every patient with at least one outpatient visit, identify if they have been diagnosed with any of the 22 conditions listed in the diseaseMap table at any time point. You will need to consider all three ICD columns in the outpatientVisit file (not just one). Create a table with the rate of disease for each condition for men, women, and all. It should look like this, where the XX% is the percent with the condition:

```{r}
# Load necessary libraries
library(dplyr)
library(readr)

# Load the datasets
outpatient_visits <- read_csv("OutpatientVisit.csv")
patients <- read_csv("Patient.csv")
clinics <- read_csv("Clinic.csv")
staff <- read_csv("Staff.csv")
mortality <- read_csv("Mortality.csv")
icd_codes <- read_csv("ICDCodes.csv")
disease_map <- read_csv("DiseaseMap.csv")

# Check structure of datasets
str(outpatient_visits)
str(patients)
str(clinics)
str(staff)
str(mortality)
str(icd_codes)
str(disease_map)

# Join the  outpatient visits with patient information to add Gender
outpatient_with_gender <- outpatient_visits %>%
  left_join(patients, by = "PatientID")

# Reshape outpatient visits to long-format (one row per ICD code)
outpatient_with_diseases <- outpatient_with_gender %>%
  pivot_longer(cols = starts_with("ICD10"), names_to = "ICD_column", values_to = "ICD_code") %>%
  # Join ICD codes to get ICD description
  left_join(icd_codes, by = c("ICD_code" = "ICD10")) %>%
  # Join disease map to get the disease condition
  left_join(disease_map, by = c("ICD_code" = "ICD10"))

# Create  list of disease to filter-by
disease_list <- disease_map$Condition

# Filter data to only include records where the disease condition is in the disease list
outpatient_with_diseases_filtered <- outpatient_with_diseases %>%
  filter(Condition %in% disease_list) %>%
  distinct(PatientID, Condition, Gender)

```

```{r}
# Calculate the disease rate by condition for men, women,for all
disease_gender_rate <- outpatient_with_diseases_filtered %>%
  group_by(Condition, Gender) %>%
  summarise(
    Total_Patients = n_distinct(PatientID),
    .groups = "drop"
  ) %>%
  # Pivot the data to get separate columns for each gender and 'All'
  pivot_wider(names_from = Gender, values_from = Total_Patients, values_fill = list(Total_Patients = 0)) %>%
  # Add a column for the total number of patients
  mutate(
    All = rowSums(select(., -Condition), na.rm = TRUE),
    Total_All = sum(All)
  ) %>%
  # Calculate the percent for each condition for each gender and 'All'
  mutate(
    Mens_Percentage = (male / Total_All) * 100,
    Womens_Percentage = (female / Total_All) * 100,
    All_Percent = (All / Total_All) * 100
  )

# Final table with the disease rates for men, women, and all
final_disease_rt <- disease_gender_rate %>%
  select(Condition, Mens_Percentage, Womens_Percentage, All_Percent)
```

```{r}
#  final result
print(final_disease_rt)
```

```{r}
test_that("summaries populated", expect_equal(nrow(final_disease_rt), 22))
```


3.Calculate the mortality rate for every year between 2005 and 2018. Is it generally increasing, or decreasing? Assume patients are only at risk of death as of their first visit (in the outpatient Visit file). Once they have died, they are no longer at risk in subsequent year.

```{r}
# Load necessary libraries
library(dplyr)
library(lubridate)

# Load the datasets
outpatient_visits <- read_csv("OutpatientVisit.csv")
mortality <- read_csv("Mortality.csv")
patients <- read_csv("Patient.csv")


#  Identify the first outpatient visit for each patient
first_visit <- outpatient_visits %>%
  group_by(PatientID) %>%
  summarise(FirstVisitDate = min(VisitDate), .groups = "drop")

# Merge with mortality data to get the date of death for each patient
mortality_data <- mortality %>%
  left_join(patients, by = "PatientID") %>%
  select(PatientID, DateOfDeath)

# Merge first visit and mortality data to determine at-risk period
patient_data <- first_visit %>%
  left_join(mortality_data, by = "PatientID") %>%
  mutate(
    DateOfDeath = as.Date(DateOfDeath),
    FirstVisitDate = as.Date(FirstVisitDate)
  )

# Step 3: Calculate mortality rates for each year between 2005 and 2018
mortality_rate_by_year <- patient_data %>%
  filter(!is.na(FirstVisitDate)) %>%
  # For each year of first visit, we need to calculate mortality rate
  mutate(YearOfFirstVisit = year(FirstVisitDate)) %>%
  filter(YearOfFirstVisit >= 2005 & YearOfFirstVisit <= 2018) %>%
  group_by(YearOfFirstVisit) %>%
  summarise(
    TotalPatients = n_distinct(PatientID),
    TotalDeaths = sum(!is.na(DateOfDeath) & year(DateOfDeath) == YearOfFirstVisit),
    MortalityRate = (TotalDeaths / TotalPatients) * 100,
    .groups = "drop"
  )

# mortality rate by the year
print(mortality_rate_by_year)

```


```{r}
#Plot the mortality rates for years to check  trends
library(ggplot2)

ggplot(mortality_rate_by_year, aes(x = YearOfFirstVisit, y = MortalityRate)) +
  geom_line() +
  geom_point() +
  ggtitle("Mortality Rate by Year (2005 - 2018)") +
  xlab("Year") +
  ylab("Mortality Rate (%)") +
  theme_minimal()
```

***INTEREPRETATION:
The death rate data from 2005 to 2018 indicates a general downward trend, beginning at 5.858% in 2005 and falling to its lowest point of 3.125% in 2017, with occasional fluctuations.

The greatest stable reduction occurred between 2008 and 2011, with rates progressively falling from 5.101% to 4.557%, indicating persistent improvements in patient care over this time.

However, the latter years (2014-2018) demonstrated higher volatility, with considerable rises in 2016 (5.897%) and 2018 (5.263%), demonstrating that, while total mortality reduced during the 13-year period, there were significant year-to-year changes in the more recent years.

```{r}
# Check the rate is increasing or decreasing
trend <- ifelse(lm(MortalityRate ~ Year, data = mortality_rates)$coefficients["Year"] > 0, 
                "increasing", "decreasing")
cat("The mortality rate is generally", trend, "\n")

```

```{r}
get_mortality_rate <- function(year) {
  rate <- mortality_rate_by_year %>%
    filter(YearOfFirstVisit == year) %>%
    pull(MortalityRate)
  if (length(rate) == 0) return(NA)  # Return NA if year not found
  return(rate)
}                                                                                                           #Test mortality rates for all years between 2005 and 2018
test_that("Mortality rates are correct for each year", {
  for (year in 2005:2018) {
    calculated_rate <- get_mortality_rate(year)
    expected_rate <- mortality_rate_by_year %>%
      filter(YearOfFirstVisit == year) %>%
      pull(MortalityRate)
    expect_equal(calculated_rate, expected_rate)  
  }
})
```

